Manuscript results: Forecasting influenza using ARIMA and ARIMAX

### 2025-06-18 manuscript_results.Rmd
update

Datasets with mean differences, standard deviations and Wilcoxon Holm adjusted p-values

@@

# Create a new column "Model_Type" based on whether "NoLg" is in the Model name
p_values_and_impr$Model_Type <- ifelse(grepl("_LB$", p_values_and_impr$Model), "LB", "no_LB")

@@

```{r}
# Define significance threshold  
alpha <- 0.05

# Get model order based on WeekAhead == 1  
model_order <- p_values_and_impr %>%
  filter(WeekAhead == 1) %>%
  arrange(desc(m)) %>%
  pull(Model)  # Extract ordered model names

# Prepare data with significance flag and ordering  
p_df <- p_values_and_impr %>%
  mutate(Significance = ifelse(p_values < alpha, "Significant", "Not Significant"),
         Model_Type = ifelse(grepl("_LB$", Model), "With log-back transformation", "Without log-back transformation"),
         Model = factor(Model, levels = model_order))  # Apply ordered factor

# Heatmap plot with models ordered by WeekAhead == 1  
plot_m_heatmap <- ggplot(p_df, aes(x = factor(WeekAhead), y = Model, fill = m)) +
  geom_tile(aes(fill = ifelse(Significance == "Significant", m, NA)), color = "black") +  # Fill only significant ones
  geom_tile(data = p_df %>% filter(Significance == "Not Significant"), 
            aes(x = factor(WeekAhead), y = Model), fill = "gray80", color = "black") +  # Gray for non-significant
  geom_text(aes(label = paste0("m:", round(m, 1),", sd:", round(sd, 1))), 
            color = "black", size = 4) +  # Add text labels
  scale_fill_gradient2(
  low = "cyan", 
  mid = "white", 
  high = "red", 
  midpoint = 0,     # <-- Sets 0 as the midpoint
  na.value = "gray80", 
  name = "Mean difference", 
  limits = c(-35, 35)       
)+

  labs(title = "Mean differences between models' WIS and the baseline across 48 states",
       subtitle = "Gray boxes indicate models that are not significantly different from the baseline (p>0.05)",
         caption = "m = mean difference, sd = standard deviation",
       x = "Target Ahead",
       y = "") + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
  plot.caption = element_text(size = 16),
    plot.title = element_text(size = 18, face = "bold"),
    plot.subtitle = element_text(size = 14),) +
  facet_wrap(~ Model_Type, scales = "free_y")  # Separate NoLg and other models

@@

model_approaches <- summary_impr %>%
  mutate(
    LogBack = grepl("_LB$", Model),
    Type = sub("_.*", "", Model),  # take everything before the first underscore
    Family = sub("^.*?_(AR|ADJ|EPI|AVG|TMP)(_LB)?$", "\\1", Model)
  )

model_approaches <- model_approaches %>%
  mutate(Family = ifelse(Family == "AR", "AAR", Family)) # rename because the intercept is in alphabetic order and I want to use the baseline

@@

Now let's plot a map of percentage of WIS improvement for each state.

```{r}
# MAP 1 week ahead

W1_map <- data.frame(
  STATE = W1$STATE,
  percentage_improvement = W1_percentage_of_improvement$AUTO_AVG_LB,
  AUTO_AVG=W1$AUTO_AVG_LB
)

# Merge the best model data with the states map
W1_map <- states %>%
  left_join(W1_map, by = c("STATE")) %>%
  drop_na()

################################### MAP 2 weeka ahead

W2_map <- data.frame(
  STATE = W2$STATE,
  percentage_improvement = W2_percentage_of_improvement$AUTO_AVG_LB,
  AUTO_AVG=W2$AUTO_AVG_LB
)

# Merge the best model data with the states map
W2_map <- states %>%
  left_join(W2_map, by = c("STATE"))%>%
  drop_na()

##################################### MAP 3 weeka ahead

W3_map <- data.frame(
  STATE = W3$STATE,
  percentage_improvement = W3_percentage_of_improvement$AUTO_AVG_LB,
  AUTO_AVG=W3$AUTO_AVG_LB
)

# Merge the best model data with the states map
W3_map <- states %>%
  left_join(W3_map, by = c("STATE"))%>%
  drop_na()

############################### MAP 4 weeka ahead

W4_map <- data.frame(
  STATE = W4$STATE,
  percentage_improvement = W4_percentage_of_improvement$AUTO_AVG_LB,
  AUTO_AVG=W4$AUTO_AVG_LB
)

# Merge the best model data with the states map
W4_map <- states %>%
  left_join(W4_map, by = c("STATE"))%>%
  drop_na()
```

Let's plot some maps with the best models in each state and the percentage of improvement compared to the AUTO ARIMA models for the same state.
Here I define some colors categories that I will use on the next maps.

@@

ES_1WEEK <- ggplot(W1_map) +
  geom_sf(aes(fill = AUTO_AVG)) +  # Fill based on the best model
  scale_fill_gradient2(low = "skyblue", mid = "lightyellow2", high = "red", midpoint = 100, limits = c(0, 352)) +  
  ggtitle("1 weeks ahead") +
  theme_light() +
  theme(legend.position = "right",
    plot.title = element_text(hjust = 0.5)) +
  geom_sf_text(data = W1_map, aes(label = round(AUTO_AVG,0)),  # Round to whole numbers
               size = 3,
               color = "black",
               check_overlap = TRUE,fontface = "bold") +  # Display percentage improvement in each state
  labs(
    fill = "mean WIS",  
       x = "",
       y = ""
  )  # Adding a subtitle

@@

Combine in a 2x2 grid

>>>>

Let create a combined dataset with mean differences, standard deviations and Wilcoxon Holm adjusted p-values

@@

# Create a new column "Model_Type" based on model with and withouth log-back transformation
p_values_and_impr$Model_Type <- ifelse(grepl("_LB$", p_values_and_impr$Model), "LB", "no_LB")

@@

# Get model order based based on lower values on WeekAhead == 1  
model_order <- p_values_and_impr %>%
  filter(WeekAhead == 1) %>%
  arrange(desc(m)) %>%
  pull(Model)

# Prepare data with significance level, ordering, and model name
p_df <- p_values_and_impr %>%
  mutate(Significance = ifelse(p_values < 0.05 , "Significant", "Not Significant"),
         Model_Type = ifelse(grepl("_LB$", Model), "With log-back transformation", "Without log-back transformation"),
         Model = factor(Model, levels = model_order))

# Heatmap plot  
plot_m_heatmap <- ggplot(p_df, aes(x = factor(WeekAhead), y = Model, fill = m)) +
  geom_tile(aes(fill = ifelse(Significance == "Significant", m, NA)), color = "black") +  # Fill only significant ones
  geom_tile(data = p_df %>% filter(Significance == "Not Significant"), 
            aes(x = factor(WeekAhead), y = Model), fill = "gray80", color = "black") +  # Gray for non-significant
  geom_text(aes(label = paste0("m:", round(m, 1),", sd:", round(sd, 1))), 
            color = "black", size = 4) +  # Add text labels
  scale_fill_gradient2(
  low = "cyan", 
  mid = "white", 
  high = "red", 
  midpoint = 0,     # <-- Sets 0 as the midpoint
  na.value = "gray80", 
  name = "Mean difference", 
  limits = c(-35, 35)       
)+
  labs(title = "Mean differences between models' WIS and the baseline across 48 states",
       subtitle = "Gray boxes indicate models that are not significantly different from the baseline (p>0.05)",
         caption = "m = mean difference, sd = standard deviation",
       x = "Target Ahead",
       y = "") + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
  plot.caption = element_text(size = 16),
    plot.title = element_text(size = 18, face = "bold"),
    plot.subtitle = element_text(size = 14),) +
  facet_wrap(~ Model_Type, scales = "free_y")

@@

```{r}
# get model approaches and mean improvement
model_approaches <- summary_impr %>%
  mutate(
    LogBack = grepl("_LB$", Model),
    Type = sub("_.*", "", Model),  # extract the model type
    Family = sub("^.*?_(AR|ADJ|EPI|AVG|TMP)(_LB)?$", "\\1", Model)
  )

# rename "AR" because the intercept is in alphabetic order and I want to use the baseline
model_approaches <- model_approaches %>%
  mutate(Family = ifelse(Family == "AR", "AAR", Family))

@@

Now let's organize the data to plot some maps with the best models.

```{r}
# MAP 1 week ahead
W1_map <- data.frame(
  STATE = W1$STATE,
  percentage_improvement = W1_percentage_of_improvement$AUTO_AVG_LB,
  AUTO_AVG=W1$AUTO_AVG_LB
)

# Include geometry
W1_map <- states %>%
  left_join(W1_map, by = c("STATE")) %>%
  drop_na()

# MAP 2 weeks a ahead
W2_map <- data.frame(
  STATE = W2$STATE,
  percentage_improvement = W2_percentage_of_improvement$AUTO_AVG_LB,
  AUTO_AVG=W2$AUTO_AVG_LB
)

# Include geometry
W2_map <- states %>%
  left_join(W2_map, by = c("STATE"))%>%
  drop_na()

# MAP 3 weeks a ahead
W3_map <- data.frame(
  STATE = W3$STATE,
  percentage_improvement = W3_percentage_of_improvement$AUTO_AVG_LB,
  AUTO_AVG=W3$AUTO_AVG_LB
)

# Include geometry
W3_map <- states %>%
  left_join(W3_map, by = c("STATE"))%>%
  drop_na()

# MAP 4 weeks ahead
W4_map <- data.frame(
  STATE = W4$STATE,
  percentage_improvement = W4_percentage_of_improvement$AUTO_AVG_LB,
  AUTO_AVG=W4$AUTO_AVG_LB
)

# Include geometry
W4_map <- states %>%
  left_join(W4_map, by = c("STATE"))%>%
  drop_na()
```

Let's plot some maps with mean WIS for each state.

@@

ES_1WEEK <- ggplot(W1_map) +
  geom_sf(aes(fill = AUTO_AVG)) +  # Fill based on the best model
  scale_fill_gradient2(low = "skyblue", mid = "lightyellow2", high = "red", midpoint = 100, limits = c(0, 352)) +  
  ggtitle("1 week ahead") +
  theme_light() +
  theme(legend.position = "right",
    plot.title = element_text(hjust = 0.5)) +
  geom_sf_text(data = W1_map, aes(label = round(AUTO_AVG,0)),  # Round to whole numbers
               size = 3,
               color = "black",
               check_overlap = TRUE,fontface = "bold") +  # Display percentage improvement in each state
  labs(
    fill = "mean WIS",  
       x = "",
       y = ""
  )  # Adding a subtitle

@@

Combining maps in a 2x2 grid
